# Prediction
## Gerald
Country: Singapore (character)
Animal: Chicken (character)
Year: 2025 (numeric)
production: 5.123 (numeric)
Pred-rows: 10000
pred-cols: 10

## Jun kiat
Country: Singapore (character)
Animal: Pig (character)
Year: 2022 (numeric)
production: 35.8 (numeric)
Pred-rows: 200
pred-cols: 20

## Adam
Country: Singapore (character)
Animal: Beef and buffalo (character)
Year: 2019 (numeric)
production: 50 (numeric)
Pred-rows: 100
pred-cols: 10

## Jia Jia
Country: Singapore (character)
Animal: Beef (character)
Year: 2018 (numeric)
production: 21.2 (numeric)
Pred-rows: 130
pred-cols: 4

## Bao Quan
Country: Singapore (character)
Animal: Pork (character)
Year: 2020 (numeric)
production: 24  (numeric)
Pred-rows: 145
pred-cols: 5


# Code

```{r}
#| label: setup-meat
#| message: false
library(tidyverse)
meat <- read_csv("meat.csv")
```

```{r}
# Part 3.1
class(meat$country)
class(meat$animal)
class(meat$year)
class(meat$production)

nrow(meat)
ncol(meat)

length(unique(meat$country)) # means there is 10 countries
length(unique(meat$animal)) # means there is 4 animals
length(unique(meat$year)) # means there is 4 years
length(unique(meat$production)) # means there is 155 unique production values

head(meat)
tail(meat)

# Explaination:
```

```{r}
# Part 3.2

# check if there is any NA values in the dataset
any(is.na(meat$production))

# Create logical mask for missing production values and list in table where prodcution is NA
missing_production <- is.na(meat$production)
meat[missing_production, c("country", "animal", "year")]

# Verifier: Check numbers of rows before and after removing missing values
nrow(meat)
nrow(meat[!is.na(meat$production), ])
nrow(meat) - nrow(meat[!is.na(meat$production), ])

# Total ASEAN production WITHOUT removing NA values
sum(meat$production)

# Total ASEAN production WITH removing NA values
sum(meat$production, na.rm = TRUE)

# Explaination:
```

```{r}
# Part 3.3

# 3.3.1
# Filter the dataset to include only rows where:
# year is 2020 and animal type is poultry
poultry_2020 <- meat[meat$year == 2020 & meat$animal == "poultry", ]

# Find the maximum poultry production value in 2020
max_poultry <- max(poultry_2020$production)

# Display the row(s) where production equals the maximum value
# This identifies the country with the highest poultry production in 2020
poultry_2020[poultry_2020$production == max_poultry, ]

# Verifier: rank countries by poultry production in 2020
poultry_2020[order(poultry_2020$production, decreasing = TRUE), ]


# 3.3.2 Second-highest poultry production in 2020

poultry_ranked <- poultry_2020[
  order(poultry_2020$production, decreasing = TRUE),
]
poultry_ranked[2, ]

# 3.3.3 Countries with more poultry than pig in 2020

# Filter the dataset to include only poultry production data for 2020
poultry_2020 <- meat[meat$year == 2020 & meat$animal == "poultry", ]

# Filter the dataset to include only pig production data for 2020
pig_2020 <- meat[meat$year == 2020 & meat$animal == "pig", ]

# Sort the poultry dataset alphabetically by country
# so the country order matches the pig dataset
poultry_2020 <- poultry_2020[order(poultry_2020$country), ]

# Sort the pig dataset alphabetically by country
pig_2020 <- pig_2020[order(pig_2020$country), ]

# Check whether both country vectors are in the same order
identical(poultry_2020$country, pig_2020$country)

# Compare poultry and pig production values country by country
# TRUE means poultry production is greater than pig production
more_poultry_than_pig <- poultry_2020$production > pig_2020$production

# Count how many countries produced more poultry than pig
sum(more_poultry_than_pig)

# Display the countries where poultry production exceeded pig production
# along with their poultry production values
poultry_2020[more_poultry_than_pig, c("country", "production")]
```

```{r}
# Part 3.4
# 3.4.1 Convert animal to factor
meat$animal_factor <- factor(meat$animal)
is.factor(meat$animal_factor)

# 3.4.2 Check levels and alphabetical order
|
levels(meat$animal_factor)
animal_rank <- min_rank(levels(meat$animal_factor))
animal_rank
identical(levels(meat$animal_factor), sort(levels(meat$animal_factor)))

# 3.4.4 Collapse factor levels

library(forcats)

meat$color <- fct_collapse(
  meat$animal_factor,
  red = c("beef and buffalo", "mutton and goat", "pig"),
  white = "poultry"
)

levels(meat$color)

# 3.4.4
# No, meat$animal_factor should not be an ordered factor because the animal categories do not have a natural ranking. They are just names of different animal groups, not values that go from low to high.
```

```{r}
# Part 3.5

# Load dplyr package for recode_values()
library(dplyr)

# Create a new column called iso3
# Map each country name to its ISO 3-letter country code

meat$iso3 <- dplyr::recode(
  meat$country,
  "Brunei Darussalam" = "BRN",
  "Cambodia" = "KHM",
  "Indonesia" = "IDN",
  "Lao PDR" = "LAO",
  "Malaysia" = "MYS",
  "Myanmar" = "MMR",
  "Philippines" = "PHL",
  "Singapore" = "SGP",
  "Thailand" = "THA",
  "Viet Nam" = "VNM",
  .default = NA
)

# Check whether the iso3 column contains any missing values
any(is.na(meat$iso3))

# Display the unique country and ISO code pairs
# to confirm there are exactly 10 valid mappings
unique(meat[, c("country", "iso3")])

# Count the number of unique country-code pairs
nrow(unique(meat[, c("country", "iso3")]))
```