---
title: "Week 03 Lab: R Practice Using ASEAN Data"
subtitle: "ASEAN Meat Production 2018–2021"
author: "Team SLATEBLUE"
date: "2026-05-12"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

# Prediction Before Running Code
**Gerald**  
Country: Singapore (character)
Animal: Chicken (character)
Year: 2025 (numeric)
production: 5.123 (numeric)
Pred-rows: 10000
pred-cols: 10

**Jun kiat**  

a. columns and classes
- country: char
- production_group: char
- animal_type: char
- year: int
- production: int

b. rows and why 
- around 160, cause 10 country * 4 types of meat * 4

c. why is current layout not tidy
- dont need ASEAN, its a sum of the years production
- need production grp so its scalable in the future, for example -> fish
- empty rows

**Adam**  
a. Columns and classes: 
- couuntry (character)
- animal (character)
- year (numeric)
- production (numeric)

b. rows and why 
- 220 as there are 11 countries, 5 meat categories, 4 years

c. why is current layout not tidy
-The first column mixes different types of information.

**Jia Jia**  
a. columns and classes
- country: character
- animal: character
- year: numeric
- production: numeric

b. rows and why 
160 rows because there are 10 ASEAN countries, 4 animal categories, and 4 years from 2018 to 2021

c. why is current layout not tidy
- because years are spread across rows or sections instead of being one year column. 
- animal types are arranged as repeated table blocks instead of being one animal column. 
- contains aggregate rows or columns such as ASEAN totals and total meat production, which are summaries rather than observations

**Bao Quan**  
a. Columns:
- country
- animal_type
- year
- production

b. rows and why 
- 4 years x 5 production x 11 countries = 176
Years are being encoded and they have separate headers rather than being a column attribute.

c. why is current layout not tidy
- Animal type is being split across multiple rows instead of being in a single var.
- Countries are also being encoded rather than being in a single var.


# Read CSV
```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(waldo)
```

# 3.2 Import the ASEAN Meat Production Data
```{r}
meat_raw <- read_xlsx(
  path = "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "X.14",
  range = "A4:L36"
)

meat_raw
```


# 3.3 Tidy the Meat Production Data
```{r}
meat_tidy <- meat_raw |>
  rename(animal = `Animal Type`) |>
  
  mutate(
    year = str_extract(as.character(animal), "^20[0-9]{2}$"),
    year = as.integer(year)
  ) |>
  fill(year) |>
  
  filter(
    animal %in% c(
      "Beef and Buffalo Meat",
      "Pig meat",
      "Mutton and Goat Meat",
      "Poultry meat"
    )
  ) |>
  
  select(-ASEAN) |>
  
  pivot_longer(
    cols = -c(year, animal),
    names_to = "country",
    values_to = "production"
  ) |>
  
  mutate(
    animal = recode(
      animal,
      "Beef and Buffalo Meat" = "beef and buffalo",
      "Pig meat" = "pig",
      "Mutton and Goat Meat" = "mutton and goat",
      "Poultry meat" = "poultry"
    ),
    production = as.numeric(production)
  ) |>
  
  select(year, animal, country, production) |>
  arrange(desc(year), animal, country)
```


```{r}
head(meat_tidy, 6)
tail(meat_tidy, 6)
```


```{r}
nrow(meat_tidy)
```

**Verifier Task:contains exactly the four product strings**
```{r}
unique(meat_tidy$animal) 
```


# 3.4 Correct Cambodia’s Mutton-and-Goat Entries
```{r}
meat_tidy_original <- meat_tidy
meat_tidy <- meat_tidy_original |>
  mutate(
    production = if_else(
      country == "Cambodia" & animal == "mutton and goat",
      NA_real_,
      production
    )
  )
```

```{r}
meat_tidy_verifier <- meat_tidy_original

meat_tidy_verifier[
  meat_tidy_verifier$country == "Cambodia" &
    meat_tidy_verifier$animal == "mutton and goat",
  "production"
] <- NA_real_

waldo::compare(meat_tidy, meat_tidy_verifier)
```

# 3.5 Checkpoint: Compare meat_tidy to Reference Data
```{r}
meat_ref <- read_csv("meat_tidy_reference.csv")

meat_tidy <- meat_tidy |>
  select(country, animal, year, production)

compare(meat_tidy, meat_ref, tolerance = 1e-6)
```

# 3.6 Tidy the Population Data
```{r}
pop_raw <- read_excel(
  "ASEAN-Statistical-Yearbook-2023.xlsx",
  sheet = "I.1",
  skip = 4,
  n_max = 10
)
```


```{r}
pop_tidy <- pop_raw |>
  pivot_longer(
    cols = -Country,
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = as.integer(year),
    population = as.numeric(population)
  ) |>
  rename(
    country = Country
  ) |>
  select(country, year, population) |>
  arrange(desc(year), country)
head(pop_tidy, 6)
```


**Verifier Task**
```{r}
pop_tidy |>
  filter(str_detect(country, "Malaysia")) |>
  distinct(country) |>
  mutate(
    n_char = str_length(country),
    has_trailing_space = str_detect(country, "\\s$")
  )
```

# 3.7 Checkpoint
```{r}
pop_ref <- read_csv(
  "population_tidy_reference.csv",
  show_col_types = FALSE
)

compare(pop_tidy, pop_ref, tolerance = 1e-6)

```

# 3.8 Join Meat Production to Population
```{r}
merged <- meat_tidy |>
  left_join(pop_tidy, by = c("country", "year")) |>
  rename(pop = population) |>
  select(country, animal, year, production, pop)
merged
```

**verifier**
```{r}
compare(
  merged |> select(country, animal, year, production),
  meat_tidy
)
```

# 3.9
```{r}
merged <- merged |>
  mutate(
    meat_per_cap = production / pop * 1000
  )

merged |>
  filter(country == "Cambodia", animal == "mutton and goat")
```

**verifier**
```{r}
merged_verifier <- merged |>
  select(country, animal, year, production, pop)

merged_verifier$meat_per_cap <- merged_verifier$production / merged_verifier$pop * 1000

waldo::compare(merged, merged_verifier, tolerance = 1e-6)
```


# 3.10
```{r}
merged_ref <- read_csv("merged_reference.csv")

compare(merged, merged_ref, tolerance = 1e-6)
```

# 3.11
```{r}
ggplot(
merged,
aes(x = year, y = meat_per_cap, color = animal)
) +
geom_line() +
geom_point() +
labs(
x = NULL,
y = "Production (kg per person)",
color = "Animal",
title = "Per-Capita Meat Production in ASEAN, 2018–2021",
caption = "Source: ASEAN Statistical Yearbook 2023"
) +
xlim(2017.9, 2021.1) +
scale_y_log10(
breaks = c(0.01, 0.1, 1, 10, 100),
labels = c("0.01", "0.1", "1", "10", "100")
) +
facet_wrap(~country) +
theme(
legend.position = "inside",
legend.position.inside = c(1, 0),
legend.justification.inside = c(1, 0.05)
)
```

